Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent

### 2023-05-27 manuscript/main.tex
tweak abstract

\title{Pangenome graph layout by HOGWILD! path-guided stochastic gradient descent}

\author{ %[Heumos, \textit{et~al}.]{
  Simon Heumos\,$^{1,2 \dagger}$,
  Andrea Guarracino\,$^{3 \dagger}$,
  Jan-Niklas Manuel Schmelzle\,$^{4,5}$,
  Jiajie Li\,$^{5}$,
  Zhiru Zhang\,$^{5}$,
  Sven Nahnsen\,$^{1,2}$,
  Pjotr Prins\,$^6$,
  and Erik Garrison\,$^{6 *}$
}

\affil{\small$^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
  $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen , Tübingen 72076, Germany \\
  $^3$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
  $^4$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
  $^5$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
  $^6$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
  $^\ast$To whom correspondence should be addressed. \\
  $^\dagger$Contributed equally.}

@@

\abstract{
    High-quality low-cost genome assemblies led to sequencing of whole populations whose genomic variation can be comparatively studied in a graphical pangenome.
    %In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
    Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.
    These nonlinear structures can introduce difficulty in downstream analyses, visualization, and interpretation.
    We propose a new layout algorithm that orders the nodes of a pangenome graph using a HOGWILD! path-guided stochastic gradient descent approach: PG-SGD.
    Our implementation demonstrates that we can efficiently compute the latent structure of gigabase-scale pangenome graphs revealing their underlying biology.
}

\section{Background}
A pangenome1 models the entire set of genomic elements of a given population. It can efficiently be encoded2 in the form of a variation graph, which embeds the linear sequences of the pangenome as paths in the graphs themselves. 
\\
In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
\\
Pangenome graphs \cite{computational2016computational}. blubb.
\\
variation graph model
\\
SGD-layout. Layout in general.
\\ Very light background.

>>>>

\title{Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent}

\author{ %[Heumos, \textit{et~al}.]{
  Simon Heumos\,$^{1,2 \dagger}$,
  Andrea Guarracino\,$^{3,6 \dagger}$,
  Jan-Niklas Manuel Schmelzle\,$^{4,5}$,
  Jiajie Li\,$^{5}$,
  Zhiru Zhang\,$^{5}$,
  Sven Nahnsen\,$^{1,2}$,
  Pjotr Prins\,$^6$,
  and Erik Garrison\,$^{6 *}$
}

\affil{\small$^1$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen 72076, Germany \\
  $^2$Biomedical Data Science, Department of Computer Science, University of Tübingen, Tübingen 72076, Germany \\
  $^3$Genomics Research Centre, Human Technopole, Milan 20157, Italy \\
  $^4$Department of Computer Engineering, School of Computation, Information and Technology (CIT), Technical University of Munich, Munich 80333, Germany \\
  $^5$School of Electrical and Computer Engineering, Cornell University, Ithaca, NY 14853, USA \\
  $^6$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN 38163, USA \\
  $^\ast$To whom correspondence should be addressed. \\
  $^\dagger$Contributed equally.}

@@

\abstract{
    \textbf{Motivation:}
	The availability of many high-quality genomes has led to the need for models that allow the study of genetic variability within entire populations.
	Pangenome graphs can represent the full diversity between multiple genomes, but they may exhibit complex structures due to common patterns of genome variation.
    These nonlinear structures introduce difficulty in downstream analyses, visualization, and interpretation.
    \textbf{Results:}
	To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
	Our approach orders the nodes of a pangenome graph by leveraging the HOGWILD! strategy and it is guided by the genomes represented in the graph.
	We show that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, exposing their biological features.
}

\section{Background}
A pangenome1 models the entire set of genomic elements of a given population. It can efficiently be encoded2 in the form of a variation graph, which embeds the linear sequences of the pangenome as paths in the graphs themselves. 
\\
In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
\\
Pangenome graphs \cite{computational2016computational}. blubb.
\\
variation graph model
\\
SGD-layout. Layout in general.
\\ Very light background.
%Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.
